---
layout: post
title: Solstice sunrise-sunset
tags: [astronomy, sun, R]
category: R
year: 2013
month: 12
day: 21
summary: Sunrise-sunset lines are calculated for the winter solstice in Halifax
description: This post shows how to determine sunrise-sunset angles on the winter solstice in Halifax, NS, Canada.
---



# Introduction

The ``sunAngle()`` function in the ``oce`` package provides a handy way to determine sunrise/sunset azimuth angles, and this is used to find alignments for the winter solstice in Halifax, NS.

# Methods

First, set up the problem; these may be the only lines that need editing for other places or times.

```{r}
t0 <- as.POSIXct("2014-12-21 10:00:00", tz="UTC") # morning of 
xy <- list(x=-63.60, y=44.65)          # centre of map (Halifax)
D <- 9                                 # map span in km
```

Next, use ``sunAngle()`` from the ``oce`` package to find the time of the sunrise and the azimuth at that moment.  Here, ``uniroot()`` is used to find the time when the altitude is zero (the sun is on the horizon), and we use a search interval that should encompass sunrise at this longitude.

```{r,message=FALSE,warning=FALSE}
library(oce)
sunrise <- uniroot(function(t)
                   sunAngle(t, lat=xy$y, lon=xy$x, useRefraction=TRUE)$altitude,
                   interval=as.numeric(t0 + 3600*c(-5,5)))$root
sunrise <- numberAsPOSIXct(sunrise)
azimuth <- 90 - sunAngle(sunrise, lat=xy$y, lon=xy$x)$azimuth
```

The span ``D`` is given in kilometres, which we convert to degrees of latitude and longitude.

```{r}
D <- D / 111                           # deg
Dlon <- D / cos(xy$y * pi / 180)
```


Now it is time to show the results. The map is done with the ``OpenStreetMap`` package.  
```{r message=FALSE, warning=FALSE}
library(OpenStreetMap)
```
```{r solstice-map,fig.height=5,fig.width=5,dpi=150}
map <- openmap(c(lat=xy$y+D/2, lon=xy$x-Dlon/2),
               c(lat=xy$y-D/2, lon=xy$x+Dlon/2),
               minNumTiles=9)
plot(map)
## Draw lines along which sunrise or sunset can be sighted.
cx <- mean(par('usr')[1:2])
cy <- mean(par('usr')[3:4])
d <- diff(par('usr')[3:4]) # scales as the map
for (o in d*seq(-1, 1, length.out=60)) {
    lines(cx+c(-1,1)*d*cos(azimuth*pi/180),
          cy+o+c(-1,1)*d*sin(azimuth*pi/180), col='red')
}
mtext(paste("Solstice sunrise at ", format(sunrise-4*3600, "%Y-%m-%d %H:%M")), font=2)
```

# Exercises

* Try altering ``t0`` to see how closely this alignment holds over time.
* As above, but setting ``t0`` to an equinox.
* Invent some sighting schemes for other times of the day, perhaps based on using
  local building (viewed at a distance computed from geometry) to check on
  noontime sun elevation.

# Resources

* R source code used here: [2013-12-21-solstice.R]({{ site.url }}/assets/2013-12-21-solstice.R).
* Jekyll source code for this blog entry: [2013-12-21-solstice.Rmd](https://raw.github.com/dankelley/dankelley.github.io/master/assets/2013-12-21-solstice.Rmd)

