---
layout: post
title: Error bars in R
tags: [R]
category: R
year: 2014
month: 3
day: 5
summary: A Monte-Carlo method for computing errors bars is presented.
description: A Monte-Carlo method for computing errors bars is presented.
---

# Introduction

Error propagation can be a fair bit of work with a calculator, but it's easy with R.  Just use R in repeated calculation with perturbed quantities, and inspect the range of results.  Two methods are shown below for inspecting the range: ``sd()`` and ``quantile()``, the latter using the fact that area under a normal distribution is 0.68 when calculated between -1 and 1 standard deviation.

# Tests


## Case 1: scale factor

In this case, the answer is simple: the A uncertainty of 0.1 turns into an uncertainty of 10 for 10A.


```{r}
set.seed(123)

n <- 500
result <- vector("double", n)
A <- 1
Au <- 0.1 # uncertainty in A
for (i in 1:n) {
    Ap <- A + Au * rnorm(n=1)
    result[i] = 10 * Ap
}
D <- 0.5 * (1 - 0.68)
r <- quantile(result, probs=c(D, 1-D))
cat("value:", mean(result), "uncertainty:", sd(result), " range:", r[1], "to", r[2], "\n")

```

                                        
## Case 2: squaring

Here, we expect an uncertainty of 20 percent.


```{r}
set.seed(123)
n <- 500
result <- vector("double", n)
A <- 1
Au <- 0.1 # uncertainty in A
for (i in 1:n) {
    Ap <- A + Au * rnorm(n=1)
    result[i] = Ap^2
}
D <- 0.5 * (1 - 0.68)
r <- quantile(result, probs=c(D, 1-D))
cat("value:", mean(result), "uncertainty:", sd(result), " range:", r[1], "to", r[2], "\n")
```

## Case 3: a nonlinear function

Here, we have a hyperbolic tangent, so the expected error bar will be trickier analytically, but of course the R method remains simple.  (Note that the uncertainty is increased to ensure a nonlinear range of hyperbolic tangent.)


```{r}
set.seed(123)
n <- 500
result <- vector("double", n)
A <- 1
Au <- 0.3 # uncertainty in A
for (i in 1:n) {
    Ap <- A + Au * rnorm(n=1)
    result[i] = tanh(Ap)
}
D <- 0.5 * (1 - 0.68)
r <- quantile(result, probs=c(D, 1-D))
cat("value:", mean(result), "uncertainty:", sd(result), " range:", r[1], "to", r[2], "\n")
```
It might be of interest to construct a histogram of the results.
```{r}
hist(result)
```

# Conclusions

The computation is a simple matter of looping over a perturbed calculation.  Here, gaussian errors are assumed, but other distributions could be used (e.g. quantities like kinetic energy that cannot be distributed in a Gaussian manner).  One should experiment with the value of ``n``, of course.

# Resources

* R source code: [2014-03-05-error-bars-in-r.R]({{ site.url }}/assets/2014-03-05-error-bars-in-r.R)

