---
layout: post
title: Using memoise to cache R values
tags: [R]
category: R
year: 2014
month: 4
day: 12
summary: Dramatic speed improvements can result from caching values with memoise().
description: Dramatic speed improvements can result from caching values with memoise().
---

# Introduction

library(memoise)

The ``memoise`` package can be very handy for caching the results of slow calculations.  In interactive work, the slowest calculations can be reading data, so that is demonstrated here.  The ``microbenchmark`` package shows timing results.

# Methods and results

## Setup

First, load the package being tested, and also a benchmarking package.

```{r}
library(memoise)
library(microbenchmark)
```

## Test conventional function

The demonstration will be for reading a CTD file.

```{r}
library(oce)
microbenchmark(d<-read.oce("/data/arctic/beaufort/2012/d201211_0002.cnv"))
```

## Memoise the function

Memoising ``read.oce()`` is simple
```{r}
r <- memoise(read.oce)
```

## Measure the speed of memoised code

```{r}
microbenchmark(d <- r("/data/arctic/beaufort/2012/d201211_0002.cnv"))
```


# Conclusions

In this example, the speedup was by a factor of about 3000.

The operation tested here is quick enough for interactive work, but this is a 1-dbar file, and the time would be increased to several seconds for raw CTD data, and increased to perhaps a half minute or so if a whole section of CTD profiles is to be read.  Using ``memoise()`` would reduce that half minute to a hundredth of a second -- easily converting an annoyingly slow operation to what feels like zero time in an interactive session.

# Resources
* [R code used in this blog item]({{ site.url }}/assets/2014-04-12-memoise.R)
* [Jekyll source code for this blog item](https://raw.github.com/dankelley/dankelley.github.io/master/assets/2014-04-12-memoise.Rmd)
