---
layout: post
title: Why I use base graphics instead of ggplot2
tags: [oce, map, R]
category: R
year: 2020
month: 9
day: 30
summary: For tasks common in my work, ggplot can be slow ... 1980s slow.
---

With base graphics, a histogram of 100 million points
```{r eval=FALSE}
d <- data.frame(x=rnorm(1e8))
system.time(hist(x))
```
takes just
```
     user  system elapsed
    0.044   0.004   0.052
```
which means that the result is ready as my pinkie finger is rising from the
'Return' key.

In contrast, if I use ggplot as in
```{r eval=FALSE}
library(ggplot2)
system.time({p<-ggplot(d) + aes(x=x) + geom_histogram();print(p)})
```
I find I have to wait *much* longer.  The results are
```
  `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
     user  system elapsed
   43.877  13.766  60.200
```
so this sub-second task has ballooned to a full minute.  Since this is not a
large dataset in my line of work, the test shows that my brand-new machine acts
more like something from the 1980s.

Don't get me wrong.  I liked the eighties.  But I don't want to go back to 
1980s computing power.

And now that I'm revealing myself to be a cranky old fart, I'll add two more
complaints about ggplot defaults:

1. The function ought to choose an appropriate binwidth, instead of saying that
it has chosen a poor default.
2. While the ggplot graph looks pretty enough for a magazine, this default
low-contrast style can be a problem for some viewers and some
printing/reproduction methods.


# References and resources

1. Jekyll source code for this blog entry: [2020-09-30-dislike-ggplot.Rmd](https://raw.github.com/dankelley/dankelley.github.io/master/assets/2020-09-30-dislike-ggplot.Rmd)


